Memes: an R interface to the MEME Suite

### 2020-11-09 manuscript.Rmd
More edits.

---
title: 'Memes: an R interface to the MEME Suite'
author:
- ORCID: 0000-0003-1000-1579
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: 
 - word_document
---

@@

Biopolymers, such as DNA and protein molecules, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers that can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function. Thus, the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. **TODO: Transition sentence RE: R/motifs** Although tools
implementing these features exist within Bioconductor, they often rely on
proprietary data types with poor interoperability with other tools. **TODO: remove this sentence?** Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline tools that provide a variety of
features for motif analysis including *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memes`, a motif analysis
package that provides a seamless interface to MEME Suite tools within R.

`memes` uses standard R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and `GRanges`. `memes` outputs also
function as inputs to other `memes` commands, allowing simple construction of motif
analysis pipelines.

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression. Lastly, `memes` also provides
simple utilities for data visualization.

`memes` is designed for maximum flexibility and ease of use to allow rapid
iteration during analysis. Here we present several examples of how `memes`
allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R landscape.

@@

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers that can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, for example,
identification of kinase substrates based on similarities in phosphorylation
site sequence, or characterizing DNA elements based on shared transcription
factor binding sequences (TODO: ref kinase motif:
https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
classify, and compare motifs represents a key step in the analysis of biological
sequences. **TODO: R/Bioconductor transition sentence**. Although tools that
implement these features exist in Bioconductor, they often rely on
proprietary data types or have poor interoperability with other tools, creating
barriers to comprehensive motif analysis within R. The recent addition of
`universalmotif` to the Bioconductor landscape partially addresss these issues
by standardizing motif formats and providing a powerful set of tools for
examining the properties of motifs. However the Bioconductor motif analysis
landscape remains incomplete \cite{universalmotif}.

The MEME Suite is a series of commandline and web-based tools that provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}. The MEME Suite family of tools provides a comprehensive
set of utilities to interrogate motif content across a broad range of biological
contexts. However, several factors limit the full potential of these tools for
use in data analysis. MEME Suite tools require carefully formatted inputs to
achieve full functionality, yet few tools exist to simplify the process of data
formatting, requiring instead that users write custom code to prepare their
data, or prepare the inputs by hand, both of which have the potential to be
error prone without rigorous testing. Furthermore, the output data from each
MEME Suite tool often have complex structures that must be parsed to extract the
full suite of information, again requiring users to write custom code for this
task. Finally, although the data-generation capabilities of the MEME Suite are
excellent, the tools lack powerful ways to visualize the results. Collectively,
these factors act as barriers to adoption, preclude deeper analysis of the
data, and limit communication of results to the scientific community.

@@

The `universalmotif` data structure is a powerful R/Bioconductor representation
for motif matrices and their associated metadata \cite{universalmotif}. However,
`universalmotif` objects are not ideal for performing bulk manipulation or
queries of metadata, nor do they support complex data structures such as nested
elements. R data.frames support nesting arbitrary data types within columns, but
are not sufficient to properly store motif matrices. In order to create an
extensible, nested data structure for tracking MEME Suite data, `memes`
leverages the power of R data.frames and the `universalmotif` format by nesting
`universalmotif` objects within data.frames. These universalmotif data.frames
store information about motifs along rows, where each column stores metadata
associated with a motif, including a `motif` column holding a `universalmotif`
representation of that motif, which can be manipulated by altering columns of
the data.frame (described in detail below). Adding one-to-many relationships within this
structure is trivial, as additional unviersalmotif data.frames can be nested
within eachother. These universalmotif data.frames form the basis for a majority
of `memes` data outputs. These structures are also valid input types to `memes`
functions, and when used as such, output data are appended as new columns to the
input data, ensuring data provenance. Finally, because the universalmotif
data.frame is a standard R data.frame, it can be manipulated using base R or
`tidyverse` functions, lowering the cognitive overhead to interfacing with these
structures.

@@

universalmotif data.frames are not limited to storing motifs from MEME Suite
data, and can additionally be used to examine or manipulate the properties of
any motif list. Using `universalmotif` functions to import motif data, the
`memes` function `as_universalmotif_dataframe()` converts any `universalmotif`
object into a `universalmotif` data.frame.  Manipulation of linked metadata
columns will result in updating the linked `universalmotif` object when calling
`update_motifs()` to retain the data.frame structure, or when calling
`as_universalmotif()` to convert the data.frame back into `universalmotif`
format. In this way, `memes` enables large-scale motif database manipulation &
analysis within the preexisting R framework.

The universalmotif data.frames generated by memes are not limited to storing
motifs from MEME Suite data. They can also be used to manipulate or store other
descriptive properties of motifs from multiple sources. `universalmotif`
provides several import functions which can convert motifs from external
databases queried from the `MotifDb` package, or from files in a variety of
non-meme file formats \cite{motifdb}. Once imported in `universalmotif` format
the `memes` function `as_universalmotif_dataframe()` converts any
`universalmotif` object into a universalmotif data.frame. Similar to MEME Suite
data stored in this format, if changes are made to the linked metadata columns,
the linked universalmotif object can be updated by calling `update_motifs()` to
retain the data.frame structure for further analysis, or by calling
`as_universalmotif()` to convert the data.frame back into universalmotif format,
discarding columns which cannot be stored within `universalmotif` format.
Therefore, this data structure empowers users to import and analyze motif lists
from a diverse set of sources for use with `memes` or simply for manipulating
motif metadata using common R tools. In this way, memes enables large-scale
motif database manipulation & analysis within the preexisting R framework.

@@

Discriminative (or "differential") motif analysis, in which motifs are discovered in one set of
sequences relative to another set, can uncover biologically relevant motifs
associated with membership to distinct categories. For example, during analysis
of multiomic data, users can identify different functional categories of genomic
regions through integration with orthogonal datasets (e.g.  categorizing
transcription factor binding sites by the presence or absence of another
factor). In these instances, motif analysis acts as a powerful tool to examine whether
differences in motif content is associated with membership of distinct
categories (e.g. motifs enriched at sites where both factors are bound vs singly
bound). Although the MEME Suite allows differential enrichment testing, it does
not inherently provide a mechanism for analyzing groups of sequences in
parallel, or performing motif analysis with any understanding of data
categories. The `memes` framework enables "data-aware" motif analysis workflows
by allowing named lists of `Biostrings` objects as input to each function. If
using `GenomicRanges`, users can split peak data on a metadata column using the
base R `split` function, and then use this result as input to `get_sequence`, which
will produce a list of `BioStrings` objects where each entry is named after the
data categories from the split column. When a list is used as input to a `memes`
function, it runs the corresponding MEME Suite tool for each object in the list.
Users can also pass the name(s) of a category to the `control` argument to
enable differential analysis of the remaining list members against the
control category sequences. In this manner, `memes` enables complex data-aware
differential motif analysis workflows using simple syntax to extend the
capabilities of the MEME Suite.

@@

Here we briefly highlight each of `memes` current features for analyzing
ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
tool, and a complete worked example using `memes` to analyze ChIP-seq data can be found
in the `memes` vignettes and the package website
([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
example, we reanalyze recent work examining the causal relationship between the
binding of the transcription factor E93 to changes in chromatin accessibility
during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
utilize ChIP-seq peak calls for E93 which have been annotated according to the
change in chromatin accessibility observed before & after E93 binding. These
data are an idiomatic example of range-based genomic data (E93 ChIP peaks)
containing additional groupings (the chromatin accessibility response following
DNA binding) whose membership may be influenced by differential motif content.
We show how `memes` syntax allows complex analysis designs, how `memes`
utilities enable deep interrogation of results, and how `memes` flexible data
structures empower users to integrate the `memes` workflow with tools offered by
other R/Bioconductor packages.

@@

```{r dec_v_inc_dreme, fig.height=10, fig.width=10, echo=F, fig.cap="\\label{fig:dec_v_inc_dreme}PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites. C,D Pearson correlation heatmaps comparing similarity of Decreasing (C) and Increasing (D) motifs."}
patchwork::wrap_plots(dreme_fig_top, dreme_fig_bottom, ncol = 1) +
  patchwork::plot_annotation(tag_levels = "A")
```
Transcription factors often bind with sequence specificity to regulate activity
of nearby genomic elements. Therefore, comparison of *de-novo* motifs with known
DNA binding motifs for transcription factors can be a key step in identifying
regulators of genomic elements. The TomTom tool is used to compare unknown
motifs against a list of known motifs to identify matches, `memes` provides the
`runTomTom` function as an interface to this tool. By passing the results from
`runDreme` into `runTomTom` and searching within a database of *Drosophila* transcription
factor motifs, we can identify candidate transcription factors which may bind
the motifs associated with increasing and decreasing chromatin accessibility.

```{r}
dreme_vs_static_tomtom <- runTomTom(dreme_vs_static, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
```
Using this approach, the results can be visualized using 
the `view_tomtom_hits()` function to visually inspect the matches
assigned by TomTom, providing a simple way to assess the quality of the match. 
A representative plot of these data is shown in \ref{fig:tomtom_example}, while
the full results can be viewed in \ref{tab:tomtom_table}. We
uncover that the motifs associated with decreasing sites are highly similar to
the E93 motif. Conversely, motifs found in increasing sites match the
transcription factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These
data support the hypothesis that E93 binding to its motif may play a larger role
in chromatin closing activity, while binding of the transcription factor br is
associated with chromatin opening.
```{r tomtom_example, echo = F, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
plot_top_hit <- function(x){
  x %>% 
    view_tomtom_hits(1)
}

@@

Discovery and matching of *de-novo* motifs is only one way to find candidate
transcription factors within target sites. Indeed, in many instances, requiring
that a motif is recovered *de-novo* is not ideal, as these approaches are
"discovery based" as opposed to hypothesis driven. Another approach, implemented
by the AME tool, is to search for known motif instances in target sequences and
test for their overrepresentation relative to a background set of sequences
\cite{ame}. The `runAme()` function is the `memes` interface to the AME tool. It
accepts a set of sequences as input an control sets, and will peform enrichment
testing using a provided motif database for occurrances of each provided motif.

@@

The `memes` universalmotif data.frame structure can be used to integrate
expression data with a motif database to remove entries for transcription
factors which are not expressed. To do so, we import a *Drosophila*
transcription factor motif database generated by the Fly Factor Survey and
convert to universalmotif data.frame format \cite{fly factor}. In this database,
the `altname` column stores the gene symbol.

```{r}
fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
  as_universalmotif_dataframe()
```
```{r, echo = FALSE}
head(fly_factor, 3)
```
Next, we import a pre-filtered list of genes expressed in a timecourse of
**Drosophila** wing development.
```{r}
wing_expressed_genes <- read.csv("data/wing_expressed_genes.csv")
```

@@

`runAme` syntax is identical to `runDreme` in that discriminative designs can be
constructed by calling list entries by name. Data can be visualized using
`ame_plot_heatmap()`, revealing that the E93 motif is strongly enriched at
Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
sites, supporting the *de-novo* discovery results \ref{fig:ame_heatmap}.
Additionally, AME detects several other transcription factor motifs which
distinguish decreasing and increasing sites, providing additional clues to
potential factors which also bind E93 ChIP peaks to affect chromatin
accessibility.
```{r}
ame_vs_static <- runAme(sequences_by_response, control = "Static", database = fly_factor_expressed)
```

```{r ame_heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="\\label{fig:ame_heatmap} A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
ame_vs_static %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste0(category, "\nAccessibility")) %>% 
  dplyr::group_by(category, motif_alt_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  ame_plot_heatmap(group = category, id = motif_alt_id) +
    labs(x = "Transcription Factor",
         y = "ChIP Peak Type")
```
### Motif matching using runFimo

A striking result from these analysis is that the E93 motif is so strongly
enriched within E93 ChIP peaks which decrease in accessibility. Significance of
motif enrichment can be driven by several factors, such as quality of the query
motif relative to the canonical motif or differences in motif number in one
group relative to other groups. These questions can be explored directly by
identifying motif occurrances in target regions and examining their properties.
FIMO allows users to match motifs in input sequences while returning information
about the quality of each match in the form of a quantitative score \cite{fimo}.

@@

Using `plyranges`, matched motifs can be joined with the metadata of the ChIP peaks with which they intersect.
```{r}
e93_fimo %<>% 
  plyranges::join_overlap_intersect(chip_results)
```

@@

    m$e93 <- e93_flyfactor
    return(m)
  } %>% 
  rev %>% 
  universalmotif::view_motifs()
```
```{r, echo=F}
n_motifs <- chip_results %>% 
  plyranges::mutate(has_match = ifelse(. %over% e93_fimo, "Match", "No Match")) %>% 
  data.frame %>% 
  ggplot(aes(e93_chromatin_response)) +
    geom_bar(aes(fill = forcats::fct_rev(has_match)), position = "fill") +
     scale_fill_manual(values = c("Match" = "firebrick",
                                  "No Match" = "Black")) +
     labs(fill = "E93 Motif Match",
          y = "Fraction of Sites",
          x = "Response to E93 binding") 
```
```{r, fig.height=5, fig.width=3, echo=F}
motif_quality <- e93_fimo %>% 
  data.frame %>% 
  ggplot(aes(e93_chromatin_response, score)) +
    geom_boxplot(aes(fill = e93_chromatin_response), notch = TRUE) +
    theme_bw() +
    labs(y = "FIMO Score",
         x = "Chromatin Response in E93 ChIP Peak") +
    theme(legend.position = "none")
```

@@

In summary, `memes` establishes a powerful motif analysis environment by
leveraging the speed and utility of the MEME Suite set of tools in conjunction
with the flexible and extensive R/Bioconductor package landscape.
```{r fimo, fig.width=7.5, fig.height=9, fig.cap="\\label{fig:fimo}A. Stacked barplot showing fraction of each chromatin response category containing E93 motif matches. B. PWMs generated from E93 motif sequences detected in each chromatin response category. C. Boxplot of FIMO Score for each E93 motif within each chromatin response category. Outliers are plotted as distinct points."}
#patchwork::wrap_plots(n_motifs, motif_quality, patchwork::guide_area(), e93_stack, widths = c(0.2, 0.3, 0.5))
patchwork::wrap_plots(n_motifs, patchwork::guide_area(), 
                      e93_stack, motif_quality, 
                      ncol = 2, widths = c(0.6, 0.4),
                      guides = 'collect') +
    patchwork::plot_annotation(tag_levels = "A")
```

>>>>

---
title: 'Memes: an R interface to the MEME Suite'
author:
- ORCID: 0000-0003-1000-1579
  name: Spencer L. Nystrom
- ORCID: 0000-0001-8226-0604
  name: Daniel J. McKay
output: 
 - word_document
---

@@

Identification of biopolymer motifs represents a key step in the analysis of
biological sequences. The MEME Suite is a widely used toolkit for comprehensive
analysis  of biopolymer motifs; however, these tools are poorly integrated
within popular analysis frameworks like the R/Bioconductor project, creating
barriers to their use. Here we present `memes`, an R package which provides a
seamless R interface to the MEME Suite. `memes` provides a novel "data aware"
interface to these tools, enabling rapid and complex discriminative motif analysis workflows.
In addition to interfacing with popular MEME Suite tools, `memes` leverages
existing R/Bioconductor data structures to store the complex, multidimensional
data returned by MEME Suite tools for rapid data access and manipulation.
Finally, `memes` provides data visualization capabilities to facilitate
communication of results. `memes` is available at
[github.com/snystrom/memes](github.com/snystrom/memes).

@@

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers that can act as interaction surfaces, create structure,
or have enzymatic activity. Identification of shared motifs across multiple
sequences can provide evidence for shared function, for example, identification
of kinase substrates based on similarities in phosphorylation site sequence, or
characterizing DNA elements based on shared transcription factor binding
sequences (TODO: ref kinase motif:
https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
classify, and compare motifs represents a key step in the analysis of biological
sequences.

The MEME Suite is a widely utilized set of tools to interrogate motif content
across a broad range of biological contexts \cite{meme-suite}. With over 22,000
citations to date, and greater than 30,000 unique users of the webserver
implementation annually, the MEME Suite has emerged as a standard tool in the
field \cite{meme-usage-report-page, google-scholar-meme}. However, several
factors limit the full potential of these tools for use in data analysis. MEME
Suite tools require carefully formatted inputs to achieve full functionality,
yet few tools exist to simplify the process of data formatting, requiring
instead that users write custom code to prepare their data, or prepare the
inputs by hand, both of which have the potential to be error prone without
rigorous testing.  Furthermore, the output data from each MEME Suite tool often
have complex structures that must be parsed to extract the full suite of
information, again requiring users to write custom code for this task. Finally,
although the data-generation capabilities of the MEME Suite are excellent, the
tools lack powerful ways to visualize the results. Collectively, these factors
act as barriers to adoption, preclude deeper analysis of the data, and limit
communication of results to the scientific community.

@@

The `universalmotif` data structure is a powerful R/Bioconductor representation
for motif matrices and their associated metadata \cite{universalmotif}. However,
`universalmotif` objects are not ideal for performing bulk manipulation or
queries of metadata, nor do they support complex data structures such as nested
elements. R data.frames support nesting arbitrary data types within columns, but
are not sufficient to properly store motif matrices. In order to create an
extensible, nested data structure for tracking MEME Suite data, `memes`
leverages the power of R data.frames and the `universalmotif` format by nesting
`universalmotif` objects within data.frames. These universalmotif data.frames
store information about motifs along rows, where each column stores metadata
associated with a motif, including a `motif` column holding a `universalmotif`
representation of that motif, which can be manipulated by altering columns of
the data.frame (described in detail below). Adding one-to-many relationships within this
structure is trivial, as additional unviersalmotif data.frames can be nested
within eachother. These universalmotif data.frames form the basis for a majority
of `memes` data outputs. These structures are also valid input types to `memes`
functions, and when used as such, output data are appended as new columns to the
input data, ensuring data provenance. Finally, because the universalmotif
data.frame is a standard R data.frame, it can be manipulated using base R or
`tidyverse` functions, making it easier to interface with these
structures.

@@

The universalmotif data.frames generated by memes are not limited to storing
motifs from MEME Suite data. They can also be used to manipulate or store other
descriptive properties of motifs from multiple sources. `universalmotif`
provides several import functions which can convert motifs from external
databases queried from the `MotifDb` package, or from files in a variety of
non-meme file formats \cite{motifdb, universalmotif}. Once imported in `universalmotif` format
the `memes` function `as_universalmotif_dataframe()` converts any
`universalmotif` object into a universalmotif data.frame. Similar to MEME Suite
data stored in this format, if changes are made to the linked metadata columns,
the linked universalmotif object can be updated by calling `update_motifs()` to
retain the data.frame structure for further analysis, or by calling
`as_universalmotif()` to convert the data.frame back into universalmotif format,
discarding columns that cannot be stored within `universalmotif` format.
Therefore, this data structure empowers users to import and analyze motif lists
from a diverse set of sources for use with `memes` or for simple manipulation of
motif metadata using common R tools. In this way, memes enables large-scale
motif database manipulation & analysis within the preexisting R framework.

@@

Discriminative (or "differential") motif analysis, in which motifs are discovered in one set of
sequences relative to another set, can uncover biologically relevant motifs
associated with membership to distinct categories. For example, during analysis
of multiomic data, users can identify different functional categories of genomic
regions through integration with orthogonal datasets (e.g. categorizing
transcription factor binding sites by the presence or absence of another
factor). Although the MEME Suite allows differential enrichment testing, it does
not inherently provide a mechanism for analyzing groups of sequences in
parallel, or for performing motif analysis with any understanding of data
categories. The `memes` framework enables "data-aware" motif analysis workflows
by allowing named lists of `Biostrings` objects as input to each function. If
using `GenomicRanges`, users can split peak data on a metadata column using the
base R `split` function, and then use this result as input to `get_sequence`, which
will produce a list of `BioStrings` objects where each entry is named after the
data categories from the split column. When a list is used as input to a `memes`
function, it runs the corresponding MEME Suite tool for each object in the list.
Users can also pass the name(s) of a category to the `control` argument to
enable differential analysis of the remaining list members against the
control category sequences. In this manner, `memes` enables complex data-aware
differential motif analysis workflows using simple syntax to extend the
capabilities of the MEME Suite.

@@

Here we briefly highlight each of `memes` current features for analyzing
ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
tool, and a worked example using `memes` to analyze ChIP-seq data can be found
in the `memes` vignettes and the package website
([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
example, we reanalyze recent work examining the causal relationship between the
binding of the transcription factor E93 to changes in chromatin accessibility
during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
utilize ChIP-seq peak calls for E93 which have been annotated according to the
change in chromatin accessibility observed before & after E93 binding. These
data are an emblematic example of range-based genomic data (E93 ChIP peaks)
containing additional groupings (the chromatin accessibility response following
DNA binding) whose membership may be influenced by differential motif content.
We show how `memes` syntax allows complex analysis designs, how `memes`
utilities enable deep interrogation of results, and how `memes` flexible data
structures empower users to integrate the `memes` workflow with tools offered by
other R/Bioconductor packages.

@@

```{r dec_v_inc_dreme, fig.height=10, fig.width=10, echo=F, fig.cap="\\label{fig:dec_v_inc_dreme}PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites. C,D Pearson correlation heatmaps comparing similarity of Decreasing (C) and Increasing (D) motifs."}
patchwork::wrap_plots(dreme_fig_top, dreme_fig_bottom, ncol = 1) +
  patchwork::plot_annotation(tag_levels = "A")
```

Transcription factors often bind with sequence specificity to regulate activity
of nearby genes. Therefore, comparison of *de-novo* motifs with known DNA
binding motifs for transcription factors can be a key step in identifying
transcriptional regulators. The TomTom tool is used to compare unknown motifs
against a list of known motifs to identify matches, `memes` provides the
`runTomTom` function as an interface to this tool. By passing the results from
`runDreme` into `runTomTom` and searching within a database of *Drosophila*
transcription factor motifs, we can identify candidate transcription factors
that may bind the motifs associated with increasing and decreasing chromatin
accessibility.

```{r}
dreme_vs_static_tomtom <- runTomTom(dreme_vs_static, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
```

Using this approach, the results can be visualized using  the
`view_tomtom_hits()` function to visually inspect the matches assigned by
TomTom, providing a simple way to assess the quality of the match.  A
representative plot of these data is shown in \ref{fig:tomtom_example}, while
the full results can be viewed in \ref{tab:tomtom_table}. This analysis revealed
that the motifs associated with decreasing sites are highly similar to the E93
motif. Conversely, motifs found in increasing sites match the transcription
factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These data support
the hypothesis that E93 binding to its motif may play a larger role in chromatin
closing activity, while binding of the transcription factor br is associated
with chromatin opening.

```{r tomtom_example, echo = F, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
plot_top_hit <- function(x){
  x %>% 
    view_tomtom_hits(1)
}

@@

Discovery and matching of *de-novo* motifs is only one way to find candidate
transcription factors within target sites. Indeed, in many instances, requiring
that a motif is recovered *de-novo* is not ideal, as these approaches are
"discovery based". Another approach, implemented
by the AME tool, is to search for known motif instances in target sequences and
test for their overrepresentation relative to a background set of sequences
\cite{ame}. The `runAme()` function is the `memes` interface to the AME tool. It
accepts a set of sequences as input and control sets, and will peform enrichment
testing using a provided motif database for occurrances of each provided motif.

@@

The `memes` universalmotif data.frame structure can be used to integrate
expression data with a motif database to remove entries for transcription
factors that are not expressed. To do so, we import a *Drosophila*
transcription factor motif database generated by the Fly Factor Survey and
convert to universalmotif data.frame format \cite{fly factor}. In this database,
the `altname` column stores the gene symbol.

```{r}
fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
  as_universalmotif_dataframe()
```

```{r, echo = FALSE}
head(fly_factor, 3)
```

Next, we import a pre-filtered list of genes expressed in a timecourse of
*Drosophila* wing development.

```{r}
wing_expressed_genes <- read.csv("data/wing_expressed_genes.csv")
```

@@

`runAme` syntax is identical to `runDreme` in that discriminative designs can be
constructed by calling list entries by name. Data can be visualized using
`ame_plot_heatmap()`, revealing that the E93 motif is strongly enriched at
Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
sites, supporting the *de-novo* discovery results \ref{fig:ame_heatmap}.
Additionally, AME detects several other transcription factor motifs that
distinguish decreasing and increasing sites, providing additional clues to
potential factors that also bind with E93 affect chromatin accessibility.

```{r}
ame_vs_static <- runAme(sequences_by_response, control = "Static", database = fly_factor_expressed)
```

```{r ame_heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="\\label{fig:ame_heatmap} A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
ame_vs_static %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste0(category, "\nAccessibility")) %>% 
  dplyr::group_by(category, motif_alt_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  ame_plot_heatmap(group = category, id = motif_alt_id) +
    labs(x = "Transcription Factor",
         y = "ChIP Peak Type")
```

### Motif matching using runFimo

A striking result from these analyses is that the E93 motif is so strongly
enriched within E93 ChIP peaks that decrease in accessibility. Significance of
motif enrichment can be driven by several factors, such as quality of the query
motif relative to the canonical motif or differences in motif number in one
group relative to other groups. These questions can be explored directly by
identifying motif occurrances in target regions and examining their properties.
FIMO allows users to match motifs in input sequences while returning information
about the quality of each match in the form of a quantitative score \cite{fimo}.

@@

Using `plyranges`, matched motifs can be joined with the metadata of the ChIP peaks with which they intersect.

```{r}
e93_fimo %<>% 
  plyranges::join_overlap_intersect(chip_results)
```

@@

    m$e93 <- e93_flyfactor
    return(m)
  } %>% 
  rev %>% 
  universalmotif::view_motifs()
```

```{r, echo=F}
n_motifs <- chip_results %>% 
  plyranges::mutate(has_match = ifelse(. %over% e93_fimo, "Match", "No Match")) %>% 
  data.frame %>% 
  ggplot(aes(e93_chromatin_response)) +
    geom_bar(aes(fill = forcats::fct_rev(has_match)), position = "fill") +
     scale_fill_manual(values = c("Match" = "firebrick",
                                  "No Match" = "Black")) +
     labs(fill = "E93 Motif Match",
          y = "Fraction of Sites",
          x = "Response to E93 binding") 
```

```{r, fig.height=5, fig.width=3, echo=F}
motif_quality <- e93_fimo %>% 
  data.frame %>% 
  ggplot(aes(e93_chromatin_response, score)) +
    geom_boxplot(aes(fill = e93_chromatin_response), notch = TRUE) +
    theme_bw() +
    labs(y = "FIMO Score",
         x = "Chromatin Response in E93 ChIP Peak") +
    theme(legend.position = "none")
```

@@

In summary, `memes` establishes a powerful motif analysis environment by
leveraging the speed and utility of the MEME Suite set of tools in conjunction
with the flexible and extensive R/Bioconductor package landscape.

```{r fimo, fig.width=7.5, fig.height=9, fig.cap="\\label{fig:fimo}A. Stacked barplot showing fraction of each chromatin response category containing E93 motif matches. B. PWMs generated from E93 motif sequences detected in each chromatin response category. C. Boxplot of FIMO Score for each E93 motif within each chromatin response category. Outliers are plotted as distinct points."}
#patchwork::wrap_plots(n_motifs, motif_quality, patchwork::guide_area(), e93_stack, widths = c(0.2, 0.3, 0.5))
patchwork::wrap_plots(n_motifs, patchwork::guide_area(), 
                      e93_stack, motif_quality, 
                      ncol = 2, widths = c(0.6, 0.4),
                      guides = 'collect') +
    patchwork::plot_annotation(tag_levels = "A")
```
